---
title: "Star Wars Music Analysis Using Spotify API"
author: "Ben Borovinsky"
date: "October 25, 2020"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<!-- styling for shiny-frame output (centers iframe) -->
```{css, echo=FALSE} 
.shiny-frame {
  margin-left: 50%;
  transform: translateX(-50%);
}
```

It's no secret that Star Wars has a massive following. What began as a space opera in 1977 has evolved into a franchise with universal recognition and fandom. Star Wars has expanded from a cult following to being a significant part of culture and media.

There are many elements that make Star Wars what it is. The iconic characters, cringe dialogue, distinct planets, and, of course, the lightsabers, are all significant pieces that make up the successful franchise. In recent years, the movies themselves have been subject to criticism, but one thing has remained consistently positive since the original trilogy: the music. John Williams has scored many iconic films, winning five academy awards and being nominated for 52 (second all-time behind Walt Disney). Williams is as synonymous with Star Wars as George Lucas, for he has scored every film in the Skywalker Saga (Episode I-IX). Like Vin Scully in the world of sports broadcasting, John Williams is the greatest film composer of all-time.

I, like many other Star Wars fans, enjoy listening to John Williams outside of the movies. The music streaming application Spotify has been my medium of choice to listen to Williams' catalogue of Star Wars music, and I've listened to a lot of his music. At the end of 2019, Spotify released a personalized report of each user's listening habits, and two of the top three I listened to the most were John Williams and the London Symphony Orchestra (the other being Frank Sinatra).

Spotify's analysis of my listening habits have encouraged me to do some analytics of my own. More specifically, I want to further analyze the music of Star Wars. While I do not have an academic background in music and I consider myself at best an amateur at data analytics, I figured this would help me learn more about two things I am very passionate about: Star Wars music and data analytics.

## Setup

The first step is to load in the libraries needed for this analysis.

```{r, warning=FALSE, message=FALSE}
library(spotifyr)
library(dplyr)
library(shiny)
library(shinyWidgets)
library(knitr)
library(tidyr)
library(ggplot2)
```

```{r include=FALSE}
options(dplyr.summarise.inform = FALSE)
```

Each library includes its own set of tools that are necessary for our analysis.

* `spotifyr`: gain access to the Spotify API
* `dplyr`: grammar and data manipulation
* `shiny`: output R code as an interactive web app
* `shinyWidgets`: more options for customizing web app
* `knitr`: structural elements to R code output
* `tidyr`: modify data for analysis
* `ggplot2`: data visualization

```{r eval=FALSE, echo=TRUE}
rm(list=ls())
graphics.off()
dirpath <- dirname(rstudioapi::getSourceEditorContext()$path)
setwd(dirpath)
```

## Getting the Data

In order to do an analysis, I need to get data. Thankfully, Spotify gives users access to its API, which makes the data acquisition relatively painless. Along with the information you would expect, the data from Spotify also includes metrics for each song to help give numerical values to the attributes of a song (this is probably how Spotify comes up with similar songs based on your listening habits or when you create a playlist); these metrics are danceability, energy, speechiness, acousticness, instrumentalness, liveness, and valence.

Note that these next several blocks are not run to save time on data retrieval from Spotify (and to keep the credentials hidden).

```{r eval=FALSE, echo=TRUE}
# SPOTIFY USER CREDENTIALS
client_id <- "###"
client_secret <- "###"

# GAIN ACCESS INTO SPOTIFY
Sys.setenv(SPOTIFY_CLIENT_ID = client_id)
Sys.setenv(SPOTIFY_CLIENT_SECRET = client_secret)

tkn <- get_spotify_access_token(client_id, client_secret)
oauth <- get_spotify_authorization_code(scope = "playlist-read-private user-top-read")
```

```{r eval=FALSE, echo=TRUE}
# GET ALL WORK FROM JOHN WILLIAMS
jw_complete <- get_artist_audio_features("john williams")

# GET ONLY WORK AVAILABLE IN US MARKET
jw_us <- jw_complete[grepl("US", jw_complete$available_markets),]

# GET STAR WARS ORIGINAL SOUNDTRACKS
jw_sw <- jw_us[grepl("Star Wars:.*Original Motion Picture Soundtrack", jw_us$album_name),]

# SELECT ONLY RELEVANT COLUMNS
jw_df <- jw_sw %>%
  select(!c(artist_id,
            album_id,
            album_type,
            album_images,
            album_release_date_precision,
            track_id,
            analysis_url,
            time_signature,
            artists,
            available_markets,
            disc_number,
            explicit,
            track_href,
            is_local,
            track_preview_url,
            type,
            track_uri,
            external_urls.spotify))

# SAVE DATA IN CSV FILE AND RDS FILE
write.csv(jw_df, "./data/john_williams_star_wars_data.csv", row.names=FALSE)
```

## Data Cleaning

```{r}
song_data <- read.csv("./data/john_williams_star_wars_data.csv", stringsAsFactors=FALSE)
```

```{r}
# ADD EPISODE
song_data <- song_data %>%
  mutate(episode = with(., case_when(
    (album_release_year == 1977) ~ 4,
    (album_release_year == 1980) ~ 5,
    (album_release_year == 1983) ~ 6,
    (album_release_year == 1999) ~ 1,
    (album_release_year == 2002) ~ 2,
    (album_release_year == 2005) ~ 3,
    (album_release_year == 2015) ~ 7,
    (album_release_year == 2017) ~ 8,
    (album_release_year == 2019) ~ 9
  )))

# GET MOVIE TITLE FROM ALBUM TITLE
song_data$movie_title <- song_data$album_name %>%
  gsub(".*[:]\\s(.*)\\s\\(.*","\\1",.)

# ADD TRILOGY
ELSE <- TRUE
song_data <- song_data %>%
  mutate(trilogy = with(., case_when(
    (episode <= 3) ~ "Prequel",
    (episode <= 6) ~ "Original",
    ELSE ~ "Sequel"
  )))

# ORGANIZE DATA
song_data <- song_data[order(song_data$album_release_year, song_data$track_number),]
rownames(song_data) <- NULL

# REORDER COLUMNS (OMIT album_release_date, key, mode)
song_data <- song_data %>%
  select(
    episode:trilogy,
    track_number,
    track_name,
    artist_name,
    album_name,
    duration_ms,
    album_release_year,
    danceability:energy,
    speechiness:tempo,
    loudness,
    key_name:key_mode
  )

categories <- c("danceability", 
                "energy", 
                "speechiness", 
                "acousticness", 
                "instrumentalness", 
                "liveness", 
                "valence")
```

I still have some more cleaning to do before I start plotting the data.

```{r}
# select specific columns
song_data <- song_data %>%
    select(
        track_name,
        movie_title,
        trilogy,
        danceability:valence
    )

song_data$movie_title <- song_data$movie_title %>%
  {factor(., levels=unique(.))}
```

## Quick Analysis

```{r}
song_data %>%
  group_by(movie_title) %>%
  summarise(count = n(),
            across(all_of(categories), ~ mean(.x)),
            .groups = "keep") %>%
  kable(format = "markdown", digits = 3, row.names = FALSE)
```

```{r}
song_data %>%
  group_by(trilogy) %>%
  summarise(count = n(),
            across(all_of(categories), ~ mean(.x)),
            .groups = "keep") %>%
  kable(format = "markdown", digits = 3, row.names = FALSE)
```

```{r}
song_data[(song_data$valence > 0.5), c("track_name", "movie_title", "trilogy", "valence")] %>%
  kable(format = "markdown", digits = 3, row.names = FALSE)
```

```{r}
plot_data <- song_data %>%
  pivot_longer(cols = all_of(categories), "name", "value") %>%
  select("movie_title", "trilogy", "track_name", "name", "value")

char_lim <- 35
plot_data$track_name <- plot_data$track_name %>%
  {ifelse(nchar(.) > char_lim, paste0(trimws(substr(., 1, char_lim-5)),"..."), .)}

plot_data[names(plot_data) != "value"] <- plot_data[names(plot_data) != "value"] %>%
  {lapply(., function(x) {factor(x, levels=unique(x))})}
```

```{r}
# colors resemble soundtrack bar color (except phantom menace because its grey)
movie_colors <- c("A New Hope" = "#2E66B1", 
                  "The Empire Strikes Back" = "#E13028",
                  "Return of the Jedi" = "#188655", 
                  "The Phantom Menace" = "#492E27", 
                  "Attack of the Clones" = "#D79E2A", 
                  "Revenge of the Sith" = "#8956A3", 
                  "The Force Awakens" = "#E9D84F", 
                  "The Last Jedi" = "#C60B36", 
                  "The Rise of Skywalker" = "#49508D")
track_colors <- c("#2E66B1", "#E13028", "#188655", 
                  "#492E27", "#D79E2A", "#8956A3", 
                  "#E9D84F", "#C60B36", "#49508D")
```

```{r eval=FALSE, include=FALSE}
radar_data <- rbind(rep(1,ncol(movie_data)),rep(0,ncol(movie_data)),movie_data)
radarchart(radar_data[,categories], axistype = 4,
           plwd = 2, plty = 1, pty = 32,
           pcol = track_colors, cglcol = "grey", cglty = 1, axislabcol = "grey", 
           caxislabels = seq(0, 1, 5), cglwd = 0.8, vlcex = 1)
legend(x=1.4, y=0.7, legend = movie_data$movie_title,
       bty = "n", lwd=1 , col=track_colors, 
       text.col = "black", cex=0.8, pt.cex=1, y.intersp = 1, seg.len = 1)

```

## Interactive Application

To learn more, see [Interactive Documents](http://rmarkdown.rstudio.com/authoring_shiny.html).

```{r, echo=FALSE}
# Define UI
ui <- fluidPage(
    
    # Application title
    #titlePanel("Star Wars X Spotify Analysis"),
    
    # Sidebar with dropdown input selectors
    sidebarLayout(
        sidebarPanel(
          
            selectizeGroupUI(
                id = "my-filters",
                inline = FALSE,
                params = list(
                    trilogy <- list(inputId = "trilogy", title = "Select Trilogy", placeholder = "select"),
                    movie <- list(inputId = "movie_title", title = "Select Movie", placeholder = "select"),
                    track <- list(inputId = "track_name", title = "Select Track", placeholder = "select")
                )
            )
            
        ),
        
        # Show a plot of the generated distribution
        mainPanel(
            
            plotOutput(
                outputId = "parallelcoordinates",
                width = "100%",
                height = "400px"
            ),
            br()
            
        )
        
    ),
    
    fluidRow(
        
        div(tableOutput("table"), style = "font-size: 90%; display: flex; justify-content: center; overflow-y: scroll; height: 200px")
        
    )
    
)
server <- function(input, output) {
    
    res_tab <- callModule(
        module = selectizeGroupServer,
        id = "my-filters",
        data = song_data,
        vars = c("trilogy", "movie_title", "track_name")
    )
        
      
    res_mod <- callModule(
        module = selectizeGroupServer,
        id = "my-filters",
        data = plot_data,
        vars = c("trilogy", "movie_title", "track_name")
    )
    
    output$parallelcoordinates <- renderPlot({
      
      plot_cond <- (length(unique(res_mod()$movie_title)) > 1) & 
        (length(unique(res_mod()$track_name)) > 5)
      plot_linetype <- if (plot_cond) {res_mod()$movie_title} else {res_mod()$track_name}
      plot_colors <- if (plot_cond) {movie_colors} else {track_colors}
      
      ggplot() +
        geom_line(aes(x=name, y=value, group=track_name, color=movie_title), 
                  data=plot_data, alpha=1.0, color="grey",
                  show.legend=FALSE) +
        geom_line(aes(x=name, y=value, group=track_name,
                      color=plot_linetype), 
                  data=res_mod(), alpha=0.7, lwd=1.5, 
                  show.legend=TRUE) +
        scale_color_manual(values=rep(
          plot_colors,
          #if (length(unique(res_mod()$movie_title))>1) {movie_colors} else {track_colors}, 
          length.out=nrow(res_mod())
        )) +
        guides(color=guide_legend(override.aes = list(size=1, alpha=1.0))) +
        theme_minimal() +
        coord_cartesian(ylim = c(0, 1)) + 
        scale_x_discrete(expand = c(0,0)) +
        labs(x="",y="strength") +
        theme(text = element_text(size = 16),
              panel.background = element_rect(fill = "transparent", color = NA),
              # plot.margin = unit(c(0.5, 13.0, 0.5, 0.5), "cm"),
              # legend.position = c(1.4, 0.5),
              legend.title = element_blank(),
              legend.text = element_text(size=10),
              #legend.key.width = unit(1, "cm"),
              axis.text = element_text(color = "black"),
              axis.text.x = element_text(angle = 25, vjust = 0.7, hjust=0.7))
        
    })
    
    output$table <- renderTable({
      
      res_tab()
        
    }, striped = TRUE, hover = TRUE)
    
}
# Run the application 
shinyApp(ui = ui, server = server, options = list(width="120%", height=650))
```

As I learn more about advanced techniques in data analysis, I would like to do more with this data (i.e. get a list of similar songs given a song input from the user). 

## Inputs and Outputs

You can embed Shiny inputs and outputs in your document. Outputs are automatically updated whenever inputs change.  This demonstrates how a standard R plot can be made interactive by wrapping it in the Shiny `renderPlot` function. The `selectInput` and `sliderInput` functions create the input widgets used to drive the plot.

```{r eruptions, echo=FALSE}
inputPanel(
  selectInput("n_breaks", label = "Number of bins:",
              choices = c(10, 20, 35, 50), selected = 20),
  
  sliderInput("bw_adjust", label = "Bandwidth adjustment:",
              min = 0.2, max = 2, value = 1, step = 0.2)
)

renderPlot({
  hist(faithful$eruptions, probability = TRUE, breaks = as.numeric(input$n_breaks),
       xlab = "Duration (minutes)", main = "Geyser eruption duration")
  
  dens <- density(faithful$eruptions, adjust = input$bw_adjust)
  lines(dens, col = "blue")
})
```

## Embedded Application

It's also possible to embed an entire Shiny application within an R Markdown document using the `shinyAppDir` function. This example embeds a Shiny application located in another directory:

```{r tabsets, echo=FALSE}
shinyAppDir(
  system.file("examples/06_tabsets", package = "shiny"),
  options = list(
    width = "100%", height = 550
  )
)
```

Note the use of the `height` parameter to determine how much vertical space the embedded application should occupy.

You can also use the `shinyApp` function to define an application inline rather then in an external directory.

In all of R code chunks above the `echo = FALSE` attribute is used. This is to prevent the R code within the chunk from rendering in the document alongside the Shiny components.



